---
output:
  html_document: default
  pdf_document: default
---
# Submission for Week 2

This file has been developed for the programming engagement for week 2 of the
Coursera Data Science module Reproducible Research
This assignment makes use of data from a personal activity monitoring device.
This device collects data at 5 minute intervals through out the day. The data 
consists of two months of data from an anonymous individual collected during
the months of October and November, 2012 and include the number of steps 
taken in 5 minute intervals each day.

First set the working directory, read the data file (a csv) and convert the date variable

```{r, echo=TRUE}

setwd("~/Datasciencecoursera/Module 5 Reproducible Research/Week 2 Programming Assignment")

raw_data <- read.csv("activity.csv")

raw_data$"date" <- as.Date(raw_data$"date", "%Y-%m-%d")

```

# What is mean total number of steps taken per day?

First some code to calculate the total number of steps taken per day, then generate the histogram

```{r, echo=TRUE}

steps_per_day <- aggregate(raw_data$steps, by = list(raw_data$date), FUN = sum)
colnames(steps_per_day) <- c("Date", "Steps_per_day")

histogram <- hist(steps_per_day$Steps_per_day, col = "red", main = "Histogram of steps taken per day", xlab = "Sum of steps per day")
```

Now the second part of the assignment: Calculate the mean and median of the steps per day

```{r, echo=TRUE}

mean_steps_per_day <- mean(steps_per_day$"Steps_per_day", na.rm=TRUE)

median_steps_per_day <- median(steps_per_day$"Steps_per_day", na.rm=TRUE)

print(mean_steps_per_day)

print(median_steps_per_day)

```

Outcome: Note median and mean are very close. This could be expected on basis of the histogram which was quite symmetric

# What is the average daily activity pattern?

Assignment: Make a time series plot (i.e. type = "l") of the 5-minute interval 
(x-axis) and the average number of steps taken, averaged across all days (y-axis)

First, compose a data frame with the data per day per 5 min interval and set the column names

```{r, echo=TRUE}

steps_per_interval <- data.frame(matrix(nrow=61, ncol=288))

names(steps_per_interval) <- raw_data[1:288,3]

```

Reformat the data to the steps_per_interval data frame, then calculate the average number of steps per interval per day.

```{r, echo=TRUE}

for (i in 1:61) {

    for (j in 1:288) {
            steps_per_interval[i,j] <- raw_data[(i-1)*288+j,1]  
    }
}

average_steps_per_interval_per_day <- data.frame(matrix(nrow=288, ncol=2))
names(average_steps_per_interval_per_day) <- c("Interval", "Average number of steps")

for (i in 1:288) {

  average_steps_per_interval_per_day[i,1] <- raw_data[i,3]
  average_steps_per_interval_per_day[i,2] <- mean(steps_per_interval[,i], na.rm = TRUE)

}

plot(average_steps_per_interval_per_day[,1], average_steps_per_interval_per_day[,2], type="l", main = "Average number of steps per interval", xlab = "Interval", ylab="Number of steps (avg)" ) 

```

Then the second part of the assignment: Calculate the interval with the maximum average number of steps

```{r, echo=TRUE}

max_average <- max(average_steps_per_interval_per_day[,2], na.rm=TRUE)

rows_with_max <- average_steps_per_interval_per_day[,2] == max_average

max_interval <- average_steps_per_interval_per_day[rows_with_max,1]

print(max_interval)

```
# Imputing Missing Values

Assignment: calculate the NA's in the data set

```{r , echo=TRUE}

number_of_complete_rows <- sum(complete.cases(raw_data))

print(c("There are ", number_of_complete_rows, " complete rows")) 
print(c("There are ", (length(raw_data[,1]) - number_of_complete_rows), "rows with NAs in it"))

```
Devise a strategy for filling in all of the missing values in the
dataset. The strategy does not need to be sophisticated. For 
example, you could use the mean/median for that day, or the mean 
for that 5-minute interval, etc. Then create a new dataset that is 
equal to the original dataset but with the missing data filled in.

The chosen strategy is to replace NA's with the average value of the 
interval over all days (with the NA's omitted of course)

```{r, echo=TRUE}

imputed_interval_step_data <- steps_per_interval

for (i in 1:61) {

    for (j in 1:288) {
            if (is.na(steps_per_interval[i,j])) {
            imputed_interval_step_data[i,j] <- average_steps_per_interval_per_day[j,2]
            }  
    }
}

# calculate the sum of steps of the imputed data set

sum_imputed_data <- data.frame(matrix(nrow=1, ncol=61))

for (i in 1:61) {
sum_imputed_data[i] <- sum(imputed_interval_step_data[i,])
}

histogram <- hist(t(sum_imputed_data), col = "red", main = "Histogram of steps taken per day with imputing data", xlab = "Sum of steps per day")
```

Calculate and report the mean and median total number of steps taken
per day. Do these values differ from the estimates from the first part
of the assignment? What is the impact of imputing missing data on the
estimates of the total daily number of steps?

```{r, echo=TRUE}

mean_total_number_of_steps <- mean(t(sum_imputed_data))

print(mean_total_number_of_steps)

median_of_total_number_of_steps <- median(t(sum_imputed_data))

print(median_of_total_number_of_steps)
```

As one can see, the imputing does not change the mean (because we added only means) but it changes the median value, which is now very close to the mean

# Are there differences in activity patterns between weekdays and weekends?

First, calculate whether the day is a weekday or weekend. Add a variable to the imputed data set, then use weekdays() to calculate the day and then a simple loop to resolve whether it is a weekday or weekend day. Lastly make the variable a factor via as.factor()

```{r, echo=TRUE}

imputed_interval_step_data$date <- unique(raw_data$date)

imputed_interval_step_data$day_of_week <- weekdays(imputed_interval_step_data$date)

for (i in 1:61) {

if (imputed_interval_step_data[i,"day_of_week"] == "zaterdag" |
  imputed_interval_step_data[i,"day_of_week"] == "zondag") {
  imputed_interval_step_data[i,"day_of_week"] <- "weekend" }
else { imputed_interval_step_data[i,"day_of_week"] <- "weekday"}  
  }

# make the variable a factor

imputed_interval_step_data$day_of_week <- as.factor(imputed_interval_step_data$day_of_week)

weekday_data <- subset(imputed_interval_step_data,imputed_interval_step_data$day_of_week == "weekday")
weekend_data <- subset(imputed_interval_step_data,imputed_interval_step_data$day_of_week == "weekend")

sum_weekday_data <- data.frame(matrix(nrow=288, ncol=2))
sum_weekend_data <- data.frame(matrix(nrow=288, ncol=2))

for (i in 1:288) {
sum_weekday_data[i,2] <- sum(weekday_data[,i])
}
sum_weekday_data[1:288,1] <- names(weekday_data[1:288])

for (i in 1:288) {
sum_weekend_data[i,2] <- sum(weekend_data[,i])
}
sum_weekend_data[1:288,1] <- names(weekend_data[1:288])

# plotting the two series in one column with two rows

par(mfcol = c(2,1))

plot(sum_weekday_data[,1], sum_weekday_data[,2], type="l", main="Steps on weekdays", xlab = "Interval", ylab="Sum of steps")

plot(sum_weekend_data[,1], sum_weekend_data[,2], type="l", main="Steps on weekends", xlab = "Interval", ylab="Sum of steps")
```
# Conclusion

The effect of weekend vs weekday is clearly visible in the data. The difference
in movements suggest the person has on office job, with lesser activity,
whereas the movement in the weekend is more steady
